Programming: What is It?
========================

I want to start today's lecture with a recommendation for a mind
blowing book: Scot Aaronson's "Quantum Computing Since Democritus."

Aaronson has a very fun style in which the gist is always conveyed
without sacrificing rigor. If you've ever wondered what the heck
quantum computing is all about and whether quantum computers really
"do all the possible calculations at the same time" its a great book
to read and you hardly need all that much physics to understand it.

But the reason I'm recommending it today is because it really gets
across the basic idea of computer science which is: imagine the set of
all tasks you can accomplish by following a finite set of
instructions, some way of recording information, and a reasonable
amount of time and space? 

There is a whole fascinating history here. Mathematicians are happy to
wrestle with various notions of infinity but programs must terminate
in a finite number of steps to produce values. It isn't too innacurate
to say that computing is just finite mathematics. 

Today we will inaugurate ourselves into the wonderful world of
computing!

I'll end today's introduction by relating one of my favorite
anectdotes from the history of mathematics.

```
    I have seen some ultrafinitists go so far as to challenge the
    existence of 2^100 as a natural number, in the sense of there
    being a series of “points” of that length. There is the obvious
    “draw the line” objection, asking where in 21, 22, 23, … , 2^100
    do we stop having “Platonistic reality”? Here this … is totally
    innocent, in that it can be easily be replaced by 100 items
    (names) separated by commas. I raised just this objection with the
    (extreme) ultrafinitist Yessenin-Volpin during a lecture of
    his. He asked me to be more specific. I then proceeded to start
    with 21 and asked him whether this is “real” or something to that
    effect. He virtually immediately said yes. Then I asked about 22,
    and he again said yes, but with a perceptible delay. Then 23, and
    yes, but with more delay. This continued for a couple of more
    times, till it was obvious how he was handling this
    objection. Sure, he was prepared to always answer yes, but he was
    going to take 2100 times as long to answer yes to 2100 then he
    would to answering 21. There is no way that I could get very far
    with this.

    Harvey M. Friedman “Philosophical Problems in Logic”

```

Working with a computer is like working with Yessenin-Volpin.

Evaluation
==========

The Simplest Expressions
------------------------

We've already seen this with Bash. Today our plan is to build up an
evaluation model for R. As with Bash, we aren't going to try to get a
perfect mental model, but we will go substantially further than we did
with Bash.

We will need to think a bit more about types as well. Let's just get
started.

```{r}
10
```

If you open this document in Rstudio you will be able to click the
little green arrow at the top of the above cell (and others like it)
and see the result of evaluating the code (and any side effects).

Who can give me a rule which explains this behavior?

Here is one:

```
Rules:
1. the characters representing a number evaluates to the number.

```

Here is another snippet.

```{r}
"Hello"
```

A set of characters between `"` characters evaluates to a string.

```
Rules:
1. the characters representing a number evaluates to the number.
2. A set of characters between `"` characters evaluates to a string.
```

What about:

```{r}
'abcdef'
```

```
Rules:
1. the characters representing a number evaluates to the number.
2. A set of characters between `"` characters evaluates to a string.
   (a) or a set of characters between `'`
```

One more string related expression:

```{r}
"this includes quotes:\"\""
```

A little tricky! When we put a backslash before a character we tell
the interpreter to treat it differently than it usually would. This is
a convention widely used throughout programming (although sometimes
you see "%" used as the escape character or some other convention). In the case
of reading strings we need to modify our rule a little bit.

```
Rules:
1. the characters representing a number evaluates to the number.
2. A set of characters between unescaped `"` characters evaluates to a string.
   (a) or a set of characters between unescaped `'`
```

Bonus points: what if you want to put a backslash into a string? Any guesses?

Algebraic Expressions
---------------------

```{r}
10+11
```
```
Rules:
1. the characters representing a number evaluates to the number.
2. A set of characters between unescaped `"` characters evaluates to a string.
   (a) or a set of characters between unescaped `'`
3. When number-representations appear in an algebraic expression the
   result is the result of applying the algebraic expression to the
   numbers produced by evaluating the character representations of the
   numbers. 
```

That might seem like a long winded way of saying what is happening,
but its useful to be precise. 

```{r}
10+11*2
```

Apparently ordinary "high school" precedence rules apply to algebraic
statements. You might think this is obvious but the most interesting
languages tend to not follow these rules for some reason.

Variables
---------

```{r}
x <- 100;
```

If we evaluate the above nothing visible happens. `<-` is called
"assignment" and it is (believe it or not) much like the operators
we've encountered above. But its primary use is not to produce a
value. It creates (or modifies, as we shall see) a _binding_. A
_binding_ is a relationship between a name and a value.

```{r}
y <- 10101
y
```

Note that we can use `<-` to modify bindings as well. 

```{r}
y <- "Hello World"
y
```

Manifestly we can store strings and numbers in the same variable. Or,
speaking more carefully, a binding doesn't care about the value to
which it points.

Here is our new set of rules.

```
Rules:
1. the characters representing a number evaluates to the number.
2. A set of characters between unescaped `"` characters evaluates to a string.
   (a) or a set of characters between unescaped `'`
3. When number-representations appear in an algebraic expression the
   result is the result of applying the algebraic expression to the
   numbers produced by evaluating the character representations of the
   numbers. 
4. And expressions like `<name> <- <expr>` evaluates `<expr>` and creates a
   binding to `<name>`.
5. `<name>` evaluates to whatever binding maps `<name>` to a
   value. Its an error to mention a name that doesn't have a binding.
```

NB: a bit of experimentation will confirm that names can contain upper
and lowercase letters and underscores but not other punctuation. There
are probably other rules but we aren't going to put too fine a point
on them.

R's Suite of Assignment Operators
---------------------------------

In many languages `=` is the assignment operator. And sure enough, one
can say:

```{r}
x = 10
x
```

In R and the expected thing happens. Indeed, you may also write:

```{r}
x <<- 10
```

And get what appears to be similar behavior. The difference between
`<-` and `=` is somewhat subtle and hinges on a thing you should
probably never do anyway. The convention is to use `<-` everywhere
unless you mean `<<-`. But we can't yet understand what `<<-` means
yet.

Collections
-----------

Collection types are very important in R. Indeed, we shall see that
almost all values are in fact collections of one kind or another. But
that comes a bit later. For now, consider:

```{r}
1:10
```

Our rule is something like:

```
Rules:
1. the characters representing a number evaluates to the number.
2. A set of characters between unescaped `"` characters evaluates to a string.
   (a) or a set of characters between unescaped `'`
3. When number-representations appear in an algebraic expression the
   result is the result of applying the algebraic expression to the
   numbers produced by evaluating the character representations of the
   numbers. 
4. And expressions like `<name> <- <expr>` evaluates `<expr>` and creates a
   binding to `<name>`.
5. `<name>` evaluates to whatever binding maps `<name>` to a
   value. Its an error to mention a name that doesn't have a binding.
6. <expr1>:<expr> produces a sequence of numbers called an array
   starting at <expr1> and ending at <expr2>.
```

In actuality the semantics of `:` are somewhat more complicated but
for our purposes this is enough.

We can also denote a collection literally like this:

```{r}
c(1,2,3)
```

A few things which would be tedious to figure out by
experimentation. Collections must be "homogeneous" which means they
must contain elements which share a type. One often calls such
homogeneous collections "arrays."

A new rule:

```
Rules:
1. the characters representing a number evaluates to the number.
2. A set of characters between unescaped `"` characters evaluates to a string.
   (a) or a set of characters between unescaped `'`
3. When number-representations appear in an algebraic expression the
   result is the result of applying the algebraic expression to the
   numbers produced by evaluating the character representations of the
   numbers. 
4. And expressions like `<name> <- <expr>` evaluates `<expr>` and creates a
   binding to `<name>`.
5. `<name>` evaluates to whatever binding maps `<name>` to a
   value. Its an error to mention a name that doesn't have a binding.
6. <expr1>:<expr> produces a sequence of numbers called an array
   starting at <expr1> and ending at <expr2>.
7. An expression like: c(<e1>,<e2>,...,<en>) produces an array of the
   values produced by evaluating <e1> through <en>.
```

Fun questions which are not that important except when they are: What
order are the `<en>` evaluated in?

Function Calls
--------------

Consider:

```{r}
x <- 10*pi;
y <- sin(x);
print(y)
```

Observation:

```
1. there are variables that are pre-bound.
2. some (most) of them are "functions".
```

New rule:

```
Rules:
1. the characters representing a number evaluates to the number.
2. A set of characters between unescaped `"` characters evaluates to a string.
   (a) or a set of characters between unescaped `'`
3. When number-representations appear in an algebraic expression the
   result is the result of applying the algebraic expression to the
   numbers produced by evaluating the character representations of the
   numbers. 
4. And expressions like `<name> <- <expr>` evaluates `<expr>` and creates a
   binding to `<name>`.
5. `<name>` evaluates to whatever binding maps `<name>` to a
   value. Its an error to mention a name that doesn't have a binding.
6. <expr1>:<expr> produces a sequence of numbers called an array
   starting at <expr1> and ending at <expr2>.
7. An expression like: c(<e1>,<e2>,...,<en>) produces an array of the
   values produced by evaluating <e1> through <en>.
8. an expression like <expr>(<a1>, <a2>, ...) is a 'function call'. It evaluates 
   to the function value. For build in functions this is just given to us.
```

Question: is `c` a function? What about `+`, `-`, etc?

Answer: They are functions. In most languages they would be operators,
but in R just about everything is a function. `+`, `-` etc are _infix_
functions and the language allows us to rely on precedence. But we can
always invoke them in the "ordinary" way like this:

```{r}
`+`(1,2) 
```

```
Rules:
1. the characters representing a number evaluates to the number.
2. A set of characters between unescaped `"` characters evaluates to a string.
   (a) or a set of characters between unescaped `'`
3. When number-representations appear in an algebraic expression the
   result is the result of applying the algebraic expression to the
   numbers produced by evaluating the character representations of the
   numbers. 
4. And expressions like `<name> <- <expr>` evaluates `<expr>` and creates a
   binding to `<name>`.
5. `<name>` evaluates to whatever binding maps `<name>` to a
   value. Its an error to mention a name that doesn't have a binding.
6. <expr1>:<expr> produces a sequence of numbers called an array
   starting at <expr1> and ending at <expr2>.
7. An expression like: c(<e1>,<e2>,...,<en>) produces an array of the
   values produced by evaluating <e1> through <en>.
8. an expression like <expr>(<a1>, <a2>, ...) is a 'function call'. It evaluates 
   to the function value. For build in functions this is just given to us.
9. An expression like `<expr>`(<a1>, <a2>, ...) is a function call but <expr> must be an infix or special operator.
```

Note that even `<-` is a function! This use of functions for
everything, along with some other properties of R, make R a very hard
language to _optimize_ but there is a lot to be said for the
uniformity.

Conditionals and Loops 
----------------------

Conditionals and loops are what makes a language "Turing Complete"
which basically means you can compute anything computable with it.

In that sense they are important. I will say rather gnomically that
conditionals and especially loops are code smells, though. If you see
a lot of them you should probably think about what you are
doing. Anyway, we must cover them for completeness's sake:

First consider:

```{r}
c(T,F)
```

Apparently this produces an array of Boolean values. Two new rules:


```
Rules:
1. the characters representing a number evaluates to the number.
2. A set of characters between unescaped `"` characters evaluates to a string.
   (a) or a set of characters between unescaped `'`
3. When number-representations appear in an algebraic expression the
   result is the result of applying the algebraic expression to the
   numbers produced by evaluating the character representations of the
   numbers. 
4. And expressions like `<name> <- <expr>` evaluates `<expr>` and creates a
   binding to `<name>`.
5. `<name>` evaluates to whatever binding maps `<name>` to a
   value. Its an error to mention a name that doesn't have a binding.
6. <expr1>:<expr> produces a sequence of numbers called an array
   starting at <expr1> and ending at <expr2>.
7. An expression like: c(<e1>,<e2>,...,<en>) produces an array of the
   values produced by evaluating <e1> through <en>.
8. an expression like <expr>(<a1>, <a2>, ...) is a 'function call'. It evaluates 
   to the function value. For build in functions this is just given to us.
9. An expression like `<expr>`(<a1>, <a2>, ...) is a function call but <expr> must be an infix or special operator.
```

### Conditionals

```{r}
x <- 10;
if(x < 100){
    "Hello"
} else {
    "Goodbye"
}
```

This _evaluates to_ "Hello". Eg, we may write:

```{r}
x <- 10;
y <- if (x < 100){
         "Hello"
     } else {
         "Goodbye"
     }
```

This will give 10. 

```
Rules:
1. the characters representing a number evaluates to the number.
2. A set of characters between unescaped `"` characters evaluates to a string.
   (a) or a set of characters between unescaped `'`
3. When number-representations appear in an algebraic expression the
   result is the result of applying the algebraic expression to the
   numbers produced by evaluating the character representations of the
   numbers. 
4. And expressions like `<name> <- <expr>` evaluates `<expr>` and creates a
   binding to `<name>`.
5. `<name>` evaluates to whatever binding maps `<name>` to a
   value. Its an error to mention a name that doesn't have a binding.
6. <expr1>:<expr> produces a sequence of numbers called an array
   starting at <expr1> and ending at <expr2>.
7. An expression like: c(<e1>,<e2>,...,<en>) produces an array of the
   values produced by evaluating <e1> through <en>.
8. an expression like <expr>(<a1>, <a2>, ...) is a 'function call'. It evaluates 
   to the function value. For build in functions this is just given to us.
9. An expression like `<expr>`(<a1>, <a2>, ...) is a function call but <expr> must be an infix or special operator.
10. if(<expr1>){ <t1>; ... <tn> } else { <f1>; ... <fn> } evalutes to <tn> if <expr1> is T, and <fn> otherwise. F, 0 count as False and most other values count as true.
```

### Loops

```{r}
x <- 0;
while (x < 10) {
    x <- x + 1;
}
x
```

```
11. while(<test>) { body0; .. bodyN } evaluates the block until <test> is F.
```

```{r}
for(x in c(1,2,3)) {
    print(x)
}
```

```
12. for(<name> in <sequence-expression>) {
        <body0>;
        ...
        <bodyN>;
    } Evaluates the block many times, with <name> bound to subsequent values of 
    <sequence-expression>.
```

Functions
---------

Now we get to the good stuff.

```
quadratic_formula <- function(a,b,c){
    r <- sqrt(b*b - 4*a*c);
    n1 <- -b + r;
    n2 <- -b - r;
    d <- 2*a;
    c(n1/d, n2/d);
}
quadratic_formula(9,100,11)
```

Observation: Creating a function is just creating a certain kind of
value. Function values are bound to variables in the same way as other
kinds of values.

So we don't need to modify our evaluation rules for variable binding
for functions.

But we need to think a little bit about what is going on inside the
"body" of the function.

```{r}
quadratic_formula <- function(a,b,c){
    r <- sqrt(b*b - 4*a*c);
    n1 <- -b + r;
    n2 <- -b - r;
    d <- 2*a;
    c(n1/d, n2/d);
}
quadratic_formula(9,100,11)
c(r, n1, n2, d)
```

What will this give? Answer: an error. We bound a series of variables
_inside_ the function but these bindings do not persist outside of
it. When the function is called it creates a new environment in which
to evaluate its body.

That environment comes pre-furnished with bindings which come from its
argument list (a, b, and c in our example), but the programmer may
extend the environment using `<-` expressions.


Our rule for function definitions is sort of complicated:

```
13.  function(<a1>, ... <an>) {
         <b1>;
         ...
         <bn>
     } creates a new function. When the function is called R creates a new environment
       and evaluates the terms <b1> through <bn> in it. The functions final 
       value is <bn>.
```

Revisiting Variable Lookup
--------------------------

Now that we have the ability to create new environments we have to
think relatively carefully about how variables are looked up. So far
we've just had a single environment: the so-called "top level". We've
added bindings to the top level using `<name> <- <expr>`
statements. But now we have multiple environments with which to
interact.

Note that we can call a function immediately, without giving it a
name. And we may use a single expression instead of a block.

```{r}
x <- 1000;
(function() x)()
```

From this example we can conclude that _if_ R finds no binding in the
enclosing environment, then it looks in the parent environment for a
value. But let's put a finer point on this idea.

```{r}
make_constant_function <- function(value){
    function(){
        value;
    }
}
always_1 <- make_constant_function(1);
```

We may construct many such functions:

```{r}
always_hi <- make_constant_function("hi");
```

```
14. Inside function environments, variables are looked up using a rule called 
"lexical scope." The nearest enclosing binding _in the text of the program_ is 
returned. Thus functions "remember" their environment, even if it would otherwise
be transient in nature.
```

Given a variable binding in an environment, we have been able to
modify it using `<-`:

```{r}
x <- 10;
x <- 11; ## This modifies the binding of x.
x
```

Consider the following example. We want a function which returns a
function which acts like a counter. We provide an initial value and
an incr and then we get back a function which counts from the
initial value by the incr on each call.

```{r}

make_counter_attempt <- function(start, incr){
    value <- start;
    function(){
        old_value <- value;
        value <- value + 1;
        old_value;
    }
}
counter_from_1 <- make_counter_attempt(1,1);
counter_from_1()
counter_from_1()
```

This returns 1 no many how many times we call it. Why?

Because `<-` creates a binding in the _inner_ function
environment. It doesn't _update_ the parent environment's
_value_. We need another operator for that: `<<-`.

Here is the correct version:

```{r}
make_counter <- function(start, incr){
    value <- start;
    function(){
        old_value <- value;
        value <<- value + 1; ## Update the old value in the parent scope.
        old_value;
    }
}
counter_from_1 <- make_counter(1,1);
counter_from_1()
counter_from_1()
```

This gives us what we want.

```
15. `<-` always creates a binding in the current environment. 
`<<-` modifies the nearest enclosing binding if it exists. 
Otherwise it creates a new one.
```

Laziness
--------

Consider:

```{r}

nope <- function(x){
    TRUE
}

nope(print("Hello Everyone"));

```

This _does not_ print "Hello Everyone" because the body of the
function doesn't "use" `x` and thus the expression

```{r}
ope <- function(x){
    x
}
ope(print("Hi"));
```

Will print `x`.

Intermission
============

Our heads are probably filled to bursting at this point. But what
we've done here has just been a more elaborate version of the
exercise we went through with bash. In Bash scripting we have a set
of relatively disjoint "environments": the filesystem, aliases, and
environment variables. It isn't simplifying things too much to say
that in Bash this is all you've got by way of environments.

In R (so far) we've just seen one _kind_ of environment: one which
just maps names to values of any kind. Regular values and functions
share this environment.

The primary difference has, so far, been that function calls create
their own environments, and nested function definitions introduce
relatively complicated variable lookup rules.

Unless you get into so-called functional programming you might get
by quite a while without having to think about these things in
detail but unfortunately, in R, because of the Tidyverse, we need to
develop a relatively good level of confidence with variable lookup
rules because the Tidyverse breaks the rules pretty liberally.

